Precision Fundamental Stellar Properties with Interpretable Machine Learning

### 2023-12-05 paper/paper2/ms.tex
Paper push

\usepackage{amsmath}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{hyperref}
% Temporary packages
\usepackage{mdframed}
\usepackage{verbatim}

@@

For the purposes of this study, we use a slice of the PHOENIX grid where the alpha abundance
is set to zero. Within this slice, we allow the full range of $\Teff, \logg, \Z$. The specific
sampling of the grid points is shown in Figure $XX$. Our approach could feasibly be extended to
include alpha abundances, but given that the \texttt{gollum} interface does not yet support the access
of spectra with nonzero alpha abundances, as well as bearing a significant increase in computational
impact, we leave this for future work.

The \texttt{blase} code takes in any spectrum as input, and outputs a cloned
version of that spectrum. However, there is a hidden product to this
procedure, the state dictionary of the model itself, stored as a \texttt{.pt} file. In this lies metadata about
where spectral lines were identified by \texttt{blase}, as well as those lines'
specific properties, namely amplitude, Gaussian shape, and Lorentzian shape.
In addition, \texttt{blase} also allows the line centers to shift slightly during
cloning, meaning that in addition to the initial line centers, we
also have access to the shifted line centers. This hidden metadata is
what formed the backbone for this study and allowed us to retrieve line-by-line data
that could then be interpolated. We input the PHOENIX grid spectra into \texttt{blase}
and accessed the state dictionaries of each clone. Once this data was retrieved from the state dictionaries
across the grid, it was then aggregated into a single \texttt{pandas} DataFrame, enabling
querying of the data by any property, be it a fundamental stellar parameter or the output
line properties.

All data and code used in this study is available online and is open source. The PHOENIX grid itself takes quite a while to download
due to server bandwidth limitations; this is something that \texttt{gollum} aims to
improve upon by implementing both caching behavior and underlying data structure compression.
\texttt{gollum} itself is available on GitHub and is pip installable, although this
method lags in features significantly compared to the bleeding edge code from GitHub.
\texttt{blase} is the main library of note here, also on GitHub, while the
materials related to this study are currently residing as a fork named \texttt{blase3D}. When polished and complete,
the fork is set to merge into \texttt{blase} itself. Links to all of these are provided in the appendix.

@@

\section{Mapping Line Parameters to Fundamental Properties}
\begin{mdframed}
    \textbf{Statement: A Bidirectonal Relation exists}

    Now that we have our list of line properties for each grid point, we need to find the link 
    between these line properties and the fundamental stellar parameters. We know that there is
    some relation between the two, but we do not even know its functional form. While we have a
    ridiculously large array of methods to choose from, we can start with a simple linear
    regression with a polynomial basis. We can quickly test the efficacy of this method by 
    running it on the line we visualized earlier.
\end{mdframed}

@@

\begin{mdframed}
    \textbf{Big equation section}
\end{mdframed}

@@

\begin{mdframed}
    \textbf{Problem: missing lines}
    - Conceivable Solution 1: treat as NaNs and deal with sparsity
    - Conceivable Solution 2: increase model complexity
    \\--------\\
    When using blas\'e, we aggregate the detected lines from all of the grid points
    into one superset of lines. However, not all of the lines in this superset are
    detected at every grid point. In order to deal with this sparsity, we ignore lines
    have $xx$ detections or less. This removes the most sparse lines, but some sparsity still
    remains. The remaining 'missing lines' are dealt with by treating them as Voigt profiles
    with amplitudes of 0, and Gaussian and Lorentzian widths equal to $xx$ and $xx$ respectively.
\end{mdframed}

@@

\section{Discussion}
\begin{mdframed}
    \textbf{Revisiting Model Assumptions}

\begin{mdframed}
    \textbf{Functional Form}
    - Describe functional form options and refinement procedures
    - List possibilities, Linreg, GPs \citep{2023ARA&A..61..329A}, NNs, etc.
    - We choose LSTSQ
    - Enumerate functional form
    - Adapting model complexity-- AIC
    \textcolor{lightgray}{\blindtext}
\end{mdframed}

>>>>

\usepackage{amsmath}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{physics}
% Temporary packages
\usepackage{mdframed}
\usepackage{verbatim}

@@

For the purposes of this study, we use a slice of the PHOENIX grid where the alpha abundance
is set to zero. Within this slice, we allow the full range of $\Teff, \logg, \Z$, leaving us with 1,959
grid points, each with its own PHOENIX spectrum. The specific domain and
sampling of the grid points are shown in Figure $XX$. Our approach could feasibly be extended to
include alpha abundances, but given that the \texttt{gollum} interface does not yet support the access
of spectra with nonzero alpha abundances, as well as bearing a significant increase in computational
impact, we leave this for future work.

The \texttt{blase} code takes in any spectrum as input, and outputs a cloned
version of that spectrum. However, there is a hidden product to this
procedure, the state dictionary of the model itself, which can be written to a file. In this lies metadata about
where spectral lines were identified by \texttt{blase}, as well as those lines'
specific properties, namely amplitude, Gaussian shape, and Lorentzian shape.
In addition, \texttt{blase} also allows the line centers to shift slightly during
cloning, meaning that in addition to the initial line centers, we
also have access to the shifted line centers. This hidden metadata is what
formed the backbone for this study and allowed us to retrieve line-by-line data
that could then be interpolated.

The ability to store these models on disk in itself is quite novel, as it offers a reduction
in storage space while also being interpretable. What we effectively do is store the
generator for the data rather than the data itself, which is a new paradigm in the 
storage of model grids such as PHOENIX.
Using this method, we input the PHOENIX grid spectra into \texttt{blase},
which wrote the models to disk, and then accessed the state dictionaries of each clone. Once this data was retrieved from the state dictionaries
across the grid, it was then aggregated into a single \texttt{pandas} DataFrame, enabling
querying of the data by any property, be it a fundamental stellar parameter or the output
line properties.

Before we proceed further, it is important to mention that all data and code used in this study is available online and is open source. The PHOENIX grid itself takes quite a while to download
due to server bandwidth limitations; this is something that \texttt{gollum} aims to
improve upon by implementing both caching behavior and underlying data structure compression.
\texttt{gollum} itself is available on GitHub and is pip installable, although this
method lags in features significantly compared to the bleeding edge code from GitHub.
\texttt{blase} is the main library of note here, also on GitHub, while the
materials related to this study are currently residing in a fork named \texttt{blase3D}. When polished and complete,
the fork is set to merge into \texttt{blase} itself. Links to all of these are provided in the appendix.

@@

\section{Mapping Line Parameters to Fundamental Properties}

Now that we have our list of line properties for each grid point, we need to find the link 
between these line properties and the fundamental stellar parameters, and we want to obtain
some relation between the two. For this study, we decided to use linear interpolation to
supersample the mapping between fundamental stellar parameters and spectral line properties.

@@

While we have discussed the process so far fairly qualitatively, we now establish the
mathematical notation for the methods that we employ in this study. Our input
fundamental line properties can be written as a vector with entries as follows:

\begin{gather}
    \theta: \mqty[\Teff, \logg, \Z]
\end{gather}

The PHOENIX grid $\bm{G}$ has $K$ grid points, indexed as $k$, and gollum
evaluates the following mapping:
\begin{gather}
    \theta_k \mapsto \bm{G}_k
\end{gather}

where $\bm{G}_k$ is the $k^{th}$ PHOENIX spectrum. All PHOENIX spectra share the same
spectral axis but differ in their flux axis.

@@

@@

\section{Discussion}

While we used interpolation to model the mapping between fundamental stellar parameters
and spectral line properties, there are also other methods that could feasibly used. Very early in the study, we considered a rather simple
solution such as polynomial linear regression, but this was quickly discarded as the regressors
had difficulty retaining detail in the dataset and scaled poorly, with the number
of terms growing as the cube of the desired degree. The scaling issue would quickly lead to the regression's design matrix
having more columns than it did rows, which would cause regression to fail due to having more degrees
of freedom than data points. Interpolation was our better solution, but others exist that we 
intend to explore, such as random forest regression, Gaussian processes, and neural networks. These methods offer an increase in the generalization of
PHOENIX's spectral features by not simply memorizing the data as an interpolator would,
with the drawback that they can be trickier to implement correctly. We also make
the assumption that every single spectral line follows a Voigt profile, which gets us 
most of the way there but does admittedly fall flat in some cases.
